---
title: "Multi-Omics Simulation"
author: "Sean Hackett"
date: "3/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This simulation will attempt to construct a very rough model of a multi-omic regulatory network. This can be used to demonstrate concepts such as:

- Measuring regulators rather than using latent variables results in conditionally-independent inference.

Factors to include a general simulation:

- transcripts:
  - regulated by proteins > metabolites 
  - modify their cognate protein
- proteins
  - regulated by cognate transcript > proteins = metabolites
- metabolites
  - This can include other chemical/physical "global effect" such as chromatin state or cellular pH. 
  - regulated by proteins and metabolites
  - modify proteins (structure or activity) >> transcripts

The connections are informed by:

- Non-cognate interactions will generally exhibit preferentially attachment and follow a power-law degree distribution.
- Effect sizes can be simulated independently of network topology.

```{r env_setup, message = FALSE}
library(dplyr)
library(ggplot2)
library(ggraph) # to plot network elements
```

```{r}
n_genes <- 20
metab_frac <- 0.3
n_metabolites <- n_genes * 0.3
n_species <- n_genes * 2 + n_metabolites # genes are doubled up because: 1 transcript + 1 protein

net_connections <- igraph::sample_pa(n_species,
                                     power = 1,
                                     m = 2)
```

```{r}
edge_counts <- igraph::degree(net_connections) %>%
  table() %>%
  {tibble::tibble(degree = names(.),
                  count = unname(.))} %>%
  dplyr::mutate_all(as.integer)

ggplot(edge_counts, aes(x = degree, y = count)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

## Assign simulated nodes to molecules

```{r}
species_init <- dplyr::bind_rows(
  tibble::tibble(gene_id = 1:n_genes,
                 species_type = "transcript"),
  tibble::tibble(gene_id = 1:n_genes,
                 species_type = "protein"),
  tibble::tibble(gene_id = NA_integer_,
                 species_type = rep("metabolite", n_metabolites))
) %>%
  dplyr::mutate(species_id = sample(1:n_species))

# edge_plausibility
edge_plausibility <- tibble::tribble(~ from_type, ~ to_type, ~ interaction_type, ~ plausibility,
                "transcript", "transcript", NA_character_, 0,
                "transcript", "protein", NA_character_, 0,
                "transcript", "metabolite", NA_character_, 0,
                "protein", "transcript", "TF", 1,
                "protein", "protein", "PPI", 0.5,
                "protein", "metabolite", "enzyme", 1,
                "metabolite", "transcript", "MDI", 0.2,
                "metabolite", "protein", "MPI", 1,
                "metabolite", "metabolite", "interconversion", 1)
                
edges_df <- igraph::as_data_frame(net_connections) %>%
  tibble::as_tibble() %>%
  dplyr::rename(from = to, to = from) %>%
  dplyr::mutate_all(as.integer) %>%
  dplyr::left_join(species_init %>%
                    dplyr::select(from = species_id, from_type = species_type),
                  by = "from") %>%
  dplyr::left_join(species_init %>%
                    dplyr::select(to = species_id, to_type = species_type),
                  by = "to") %>%
  dplyr::left_join(edge_plausibility, by = c("from_type", "to_type")) %>%
  # cull edges based on plausibility
  dplyr::filter(plausibility != 0) %>%
  dplyr::group_by(from, interaction_type) %>%
  dplyr::mutate(plausible = sample(c(TRUE, FALSE), size = 1, prob = c(plausibility[1], 1-plausibility[1]))) %>%
  dplyr::ungroup() %>%
  dplyr::filter(plausible) %>%
  dplyr::select(from, to, interaction_type) %>%
  # add cognate edges
  dplyr::bind_rows(
    species_init %>%
      dplyr::filter(species_type != "metabolite") %>%
      tidyr::spread(species_type, species_id) %>%
      dplyr::select(from = transcript, to = protein) %>%
      dplyr::mutate(interaction_type = "cognate")
  )
  
species_df = species_init %>%
  dplyr::select(species_id, species_type) %>%
  dplyr::arrange(species_id)

pruned_network <- igraph::graph_from_data_frame(edges_df,
                                                directed = TRUE,
                                                vertices = species_df)
```

```{r}
edges_df %>%
  dplyr::count(interaction_type) %>%
  knitr::kable() %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

```{r}
work_with_top_cluster <- TRUE
if (work_with_top_cluster) {
  network_clusters <- igraph::clusters(pruned_network)
  isolated_vertices <- names(network_clusters$membership)[network_clusters$membership != which.max(network_clusters$csize)]
  
  pruned_network <- igraph::delete_vertices(pruned_network, isolated_vertices)
}

ggraph(pruned_network, layout = "kk") +
  geom_edge_link(arrow = arrow(type = "closed", length = unit(0.07, "inches")), color = "gray50", alpha = 0.5) +
  geom_node_point(aes(color = species_type)) +
  scale_color_brewer("Species Type", palette = "Set2") +
  theme_void() +
  theme(legend.position = "bottom")
```

## Simulating a dynamic dataset

```{r}
library(deSolve)

reffect <- function (n, mean = 0.5, shape = 10) {
  magnitudes <- rgamma(n, shape = shape, scale = mean / shape)
  effect <- magnitudes * (rbinom(n, 1, 0.5) - 0.5)*2
}

Deffects <- function(time, states, pars, species_df, basal_prod = 0.2, loss_rate = 0.2) {
  
  reg_states <- tibble::tibble(value = states) %>%
    dplyr::mutate(species_id = 1:n()) %>%
    dplyr::left_join(species_df, by = "species_id")
  
  modulated <- pars %>%
    dplyr::left_join(reg_states, by = c("from" = "species_id")) %>%
    dplyr::group_by(to) %>%
    dplyr::summarize(net_effect = sum(value * effect))
  
  reg_states <- reg_states %>%
    dplyr::left_join(modulated, by = c("species_id" = "to")) %>%
    dplyr::mutate(net_effect = ifelse(is.na(net_effect), 0, net_effect),
                  # basal production
                  basal_synthesis = ifelse(species_type == "protein", 0, basal_prod),
                  loss = -1 * value * loss_rate,
                  deriv = net_effect + basal_synthesis + loss)
  
  return (as.list(reg_states$deriv))
  
}

states <- rep(1, nrow(species_df))

pars <- edges_df %>%
  dplyr::mutate(effect = reffect(n()),
                effect = ifelse(interaction_type == "cognate", abs(effect), effect))

ggplot(pars, aes(x = effect, fill = interaction_type)) +
  geom_histogram(binwidth = 0.1)

ode(y = rep(1, nrow(species_df)),
    times = seq(0, 20, by = 0.1),
    func = Deffects,
    pars,
    species_df = species_df)


```
